Data Specialization for a Staged Lambda Calculus

### 2015-02-27 paper/splitting.tex
fix to splitting rule
Complete pass through 4.0

\textbf{\bbtwo\ term splitting.}  For any $\coltwo e A$ which reduces
to a residual $q$ (via $\diatwo e q$) and then a final value $v$ (via
$\reduce q v$), splitting $e$ (via the new relation~$\splittwosym$)
yields a pair of monostaged terms $p$ (containing all first stage
computations in $e$) and $l.r$ (containing all second stage
computations in $e$). Executing $p$ in a first pass produces $b$ (via
$\reduce p b$). Then, in the second pass, $\reduce {[b/l]r} {v'} = v$.
Since execution of the first pass serves to generate input for the
second pass, we say $p$ is a {\em precomputation} that produces a {\em
  boundary value} ($b$) for the {\em resumer} ($l.r$).

@@

% (and the relationship of these
% operations to the outputs of direct \lang\ evaluation) is provided at
% right in \ref{fig:splittingSummary}.

% $\sepredtwosym$ is defined
% by a single rule, which evaluates the precomputation and substitutes
% the resulting value in for $l$ in the resumer:
%\[
%\infer{{[b/l]r}\equiv q} {\reduce p b}
%\]

\textbf{\bbonem\ term splitting.}  For any $\colmix e A$ which reduces
to $\rtab \xi v$ (via $\redonesym$), masking yields the masked value
$\mval i q$ (via~$\vsplito$).  Then $e$ splits (via the new
relation~$\splitonesym$) into the monostage terms $\pipeM c l r$ where
$\reduce c (i,b)$, and $[b/l]r$ is the same as $q$ up to an
equivalence.  The term $c$ generated by splitting (called the {\em
  combined term}) is a monostage term encoding all the first-stage
computations in $e$, and $l.r$ (again called the {\em resumer}) is a
monostage term encoding all the second-stage computations in $e$.  The
process of splitting \bbonem\ terms is summarized at left in
\ref{fig:splittingSummary}.

%Defining the requirements for
%splitting \bbonem\ terms is challenging because unlike \bbtwo\ terms
%(which reduce to monostage residuals that can be directly compared to
%resumers), \bbonem\ terms reduce to residual tables and partial values
%that represent computations at both stages.

@@

% We define the requirements for splitting \bbonem\ terms using the
% results of masking.

Note that since $\coltwo e A$ reduces to an entirely second-stage
residual, its first-stage computations only exist to generate input
for second stage computations.  In contrast, since $\colmix e A$
reduces to a multi-stage residual table and partial value, its first
stage computations serve to both generate the value used as input for
the second stage {\em and also} produce a first-stage value.
Correspondingly, the precomputation $p$ resulting from splitting
\bbtwo\ terms reduces only to a boundary value $b$ which is passed to
the resumer; but the combined term $c$ resulting from splitting
\bbonem\ terms reduces to a tuple containing both the first stage {\em
  immediate result} $i$ and the boundary value $b$ (motivating the
name {\em combined term}).

The following subsections formally define the masking transformation
($\vsplito$), then describe the algorithm for splitting both
\bbonem\ terms ($\splitonesym$) and \bbtwo\ terms ($\splittwosym$).

>>>>

\textbf{\bbtwo\ term splitting.}  For any $\coltwo e A$ which reduces
to a residual $q$ (via $\diatwo e q$) and then a final value $v$ (via
$\reduce q v$), splitting $e$ (via the new relation~$\splittwosym$)
yields a pair of monostaged terms $p$ and $l.r$. ($p$ and $l.r$
contain all first and second stage computations in $e$ respectively.)
Executing $p$ in a first pass produces $b$ (via $\reduce p b$). Then,
in the second pass, $\reduce {[b/l]r} {v'}$, where $v'$ is the same as
$v$.  Since execution of the first pass serves to generate input for
the second pass, we say $p$ is a {\em precomputation} that produces a
{\em boundary value} ($b$) for the {\em resumer} ($l.r$).

@@

%masking yields the masked value $\mval i q$ (via~$\vsplito$)

\textbf{\bbonem\ term splitting.}  For any $\colmix e A$ which reduces
to $\rtab \xi v$ (via $\redonesym$) , splitting $e$ (via the new
relation~$\splitonesym$) again yields the pair of monostage terms $c$
and $l.r$
%(corresponding to the first stage and second stage
%computations in $e$).
Evaluating $c$ in the first pass produces the pair $(i,b)$, where the
first element $i$ matches the first stage results contained in $v$.
For example, in the simple case where $\rtab \xi v$ contains entirely
first stage computation, such as:
\begin{lstlisting}
2`[]`1` grnd{1+2}`
\end{lstlisting}
the first stage result is the value 3.  The second element of the pair
is used as input to the resumer in second pass evaluation as $\reduce
{[b/l]r} {v'}$, where $v'$ is the same as result produced by second
stage computations in $e$.  For example, in the case where $\rtab \xi
v$ contains entirely second stage computation, such as:
\begin{lstlisting}
2`[yhat |-> 2+3]`1` next{`2`yhat`1`}`
\end{lstlisting}
the second stage result is the value 5. Since in general residual
tables and partial values describe a mixture of both first and second
stage computations, in the subsequent subsection we define an
operation, called {\em masking}, that separates residual tables and
partial values into stage one and stage two results. We then describe
the algorithm for splitting both \bbonem\ terms ($\splitonesym$) and
\bbtwo\ terms ($\splittwosym$).

%Since $c$ ensapulates computations that produce both $e$'s first-stage
%result and also the boundary value required by the resumer, we refer
%to it as a {\em combined term}.

%Note that since $\coltwo e A$ reduces to an entirely second-stage
%residual, its first-stage computations only exist to generate input
%for second stage computations.  In contrast, since $\colmix e A$
%reduces to a multi-stage residual table and partial value, its first
%stage computations serve to both generate the value used as input for
%the second stage {\em and also} produce a first-stage value.
%Correspondingly, the precomputation $p$ resulting from splitting
%\bbtwo\ terms reduces only to a boundary value $b$ which is passed to
%the resumer; but the combined term $c$ resulting from splitting
%\bbonem\ terms reduces to a tuple containing both the first stage {\em
%  immediate result} $i$ and the boundary value $b$ (motivating the
%name {\em combined term}).

@@
